### 2014-03-15 thesis-proposal/parsing.tex
removed stuff

% !TEX root = omar-thesis-proposal.tex
\lstset{tabsize=2, 
basicstyle=\ttfamily\fontsize{8pt}{1em}\selectfont, 
commentstyle=\itshape\ttfamily, 
stringstyle=\ttfamily,
numbers=left, numberstyle=\scriptsize\color{gray}\ttfamily, language=ML,moredelim=[il][\sffamily]{?},mathescape=true,showspaces=false,showstringspaces=false,xleftmargin=15pt, morekeywords=[1]{tyfam,opfam,let,fn,val,def,casetype,objtype,metadata,of,*,datatype,new,valAST},deletekeywords={for,double,in},classoffset=0,belowskip=\smallskipamount,
moredelim=**[is][\color{cyan}]{SSTR}{ESTR},
moredelim=**[is][\color{OliveGreen}]{SHTML}{EHTML},
moredelim=**[is][\color{purple}]{SCSS}{ECSS},
moredelim=**[is][\color{brown}]{SSQL}{ESQL},
moredelim=**[is][\color{orange}]{SCOLOR}{ECOLOR},
moredelim=**[is][\color{magenta}]{SPCT}{EPCT}, 
moredelim=**[is][\color{gray}]{SNAT}{ENDNAT}, 
moredelim=**[is][\color{blue}]{SURL}{EURL},
moredelim=**[is][\color{SeaGreen}]{SQT}{EQT},
moredelim=**[is][\color{Periwinkle}]{SGRM}{EGRM},
moredelim=**[is][\color{YellowGreen}]{SID}{EID},
moredelim=**[is][\color{Sepia}]{SUS}{EUS}
}
\lstloadlanguages{Java,VBScript,XML,HTML}
\let\li\lstinline

\section{Type-Specific Languages}\label{aparsing}
General-purpose abstraction mechanisms like those built into modern languages are semantically quite flexible. By using a general-purpose abstraction mechanism to define an abstraction, one immediately benefits from a body of useful operations, established reasoning principles, well-optimized implementations and tool support. For example, lists can be encoded using a more general mechanism for defining  inductive datatypes. Intuitively, a list can either be empty, or be broken down into a \emph{head} element and a \emph{tail}, another list. In an ML-like language, the polymorphic list type is declared:
\begin{lstlisting}[numbers=none]
datatype 'a list = Nil | Cons of 'a * 'a list
\end{lstlisting}
By encoding lists in this way, we can immediately reason about them by structural induction and examine them by pattern matching. %The programmer only needs to provide an encoding of the structure of lists; the semantics and implementation are filled in by the general-purpose abstraction mechanism.

While these operations and reasoning principles can be quite useful, the associated general-purpose syntax can sometimes be a liability. For example, few would claim that writing a list of numbers as a sequence of \li{Cons} cells is convenient:
\begin{lstlisting}[numbers=none]
Cons(1, Cons(2, Cons(3, Cons(4, Nil))))
\end{lstlisting}

Because lists are a common data structure, many languages include \emph{literal syntax} for introducing them, e.g. \li{[1, 2, 3, 4]}. This notation is semantically equivalent to the general-purpose notation shown above, but brings cognitive benefits by drawing attention to the content of the list, rather than the nature of the encoding. To use terminology from the literature on the cognitive dimensions of notations \cite{green1996usability}, it is more \emph{terse}, \emph{visible} and \emph{maps more closely} to the intuitive notion of a list.

Although list, number and string literals are nearly ubiquitous features of modern languages, some languages also  provide specialized notation for other common data structures (like maps and sets), data formats (like XML and JSON), query languages (like regular expressions and SQL), markup languages (like HTML) and many other types of data. For example, a language with built-in notation for HTML and SQL, with type-safe \emph{interpolation} of host language terms within curly braces, might define:
\begin{lstlisting}
let webpage : HTML = SHTML<html><body><h1>Results for {EHTMLkeywordSHTML}</h1>
  <ul id="results">{EHTMLto_list_items(query(db, 
    SSQLSELECT title, snippet FROM products WHERE {ESQLkeywordSSQL} in titleESQL)SHTML}
  </ul></body></html>EHTML
\end{lstlisting}
as shorthand for:
\begin{lstlisting}
let webpage : HTML = HTMLElement(Dict.empty(), [BodyElement(Dict.empty(),
  [H1Element(Dict.empty(), [TextNode(concat($\texttt{"}$SSTRResults for $\texttt{"}$ESTR, keyword))]), 
  ULElement((Dict.add Dict.empty() ($\texttt{"}$SSTRid$\texttt{"}$ESTR, $\texttt{"}$SSTRresults$\texttt{"}$ESTR)), to_list_items(query(db, 
    SelectStmt([$\texttt{"}$SSTRtitle$\texttt{"}$ESTR, $\texttt{"}$SSTRsnippet$\texttt{"}$ESTR], $\texttt{"}$SSTRproducts$\texttt{"}$ESTR, 
      [WhereClause(InPredicate(StringLit(keyword), $\texttt{"}$SSTRtitle$\texttt{"}$ESTR))]))))]]]
\end{lstlisting}

When a specialized notation like this is not available, but the equivalent general-purpose notation is too cognitively demanding for comfort, developers typically turn to run-time mechanisms to make constructing data structures more convenient. Among the most common strategies in these situations is to simply use a string representation that is parsed at run-time. Developers across language paradigms frequently write examples like the above as:
\begin{lstlisting}
let webpage : HTML = parse_html($\texttt{"}$SSTR<html><body><h1>Results for ESTR$\texttt{"}$ + keyword + $\texttt{"}$SSTR</h1>
  <ul id=\$\texttt{\color{cyan}"}$results\$\texttt{\color{cyan}"}$>$\texttt{"}$ESTR + to_string(to_list_items(query(db, parse_sql(
  	$\texttt{"}$SSTRSELECT title, snippet FROM products WHERE '$\texttt{" + keyword + "}$' in title$\texttt{"}$ESTR)))) + 
  $\texttt{"}$SSTR</ul></body></html>$\texttt{")}$
\end{lstlisting}

Though recovering much of the notational convenience of the literal version, it is still more awkward to write, requiring explicit conversions to and from structured representations (\li{parse_html} and \li{to_string}, respectively) and escaping when the syntax of the language interferes with the syntax of string literals (line 2). Code like this also causes a number of problems beyond cognitive load. Because parsing occurs at run-time, syntax errors will not be discovered statically, causing potential problems in production scenarios. Run-time parsing also incurs performance overhead, particularly relevant when code like this is executed often (as on a heavily-trafficked website). But the most serious issue with this code is that it is fundamentally insecure: it is vulnerable to cross-site scripting attacks (line 1) and SQL injection attacks (line 3). For example, if a user entered the keyword \li{'; DROP TABLE products --}, the entire product database could be erased. These attack vectors are considered to be two of the most serious security threats on the web today \cite{owasp2013}. Although developers are cautioned to sanitize their input, it can be difficult to verify that this was done correctly throughout a codebase. The most straightforward way to avoid these problems today is to insist on structured representations, despite their inconvenience.

Unfortunately, it does not appear that this is working. Situations like this, where maintaining strong correctness, performance and security guarantees entails significant syntactic overhead, causing developers to turn to worse solutions that are more convenient, are quite common. To emphasize this, let us return to our running example of pattern literals. A small regular expression like \verb!(\d\d):(\d\d)\w*((am)|(pm))! might be written using general-purpose notation as:
\begin{lstlisting}
Seq(Group(Seq(Digit, Digit), Seq(Char($\texttt{"}$SSTR:ESTR$\texttt{"}$), Seq(Group(Seq(Digit, Digit)), 
  Seq(ZeroOrMore(Whitespace), Group(Or(Group(Seq(Char($\texttt{"}$SSTRaESTR$\texttt{"}$), Char($\texttt{"}$SSTRmESTR$\texttt{"}$))), 
  Group(Seq(Char($\texttt{"}$SSTRpESTR$\texttt{"}$), Char($\texttt{"}$SSTRmESTR$\texttt{"}$))))))))))
\end{lstlisting}
This is clearly more cognitively demanding, both when authoring the regular expression and when reading it. Among the most common strategies in these situations, for users of both object-oriented and functional languages, is to simply use a string representation that is parsed at run-time.
\begin{lstlisting}
rx_from_str($\texttt{"}$SSTR(\\d\\d):(\\d\\d)\\w*((am)|(pm))ESTR$\texttt{"}$)
\end{lstlisting}
%%
%%For example, in languages without SQL literals, developers can implement a builder pattern:
%%\begin{lstlisting}
%%new SQLQuery().SELECT("*").FROM("table").WHERE("username").Eq(username)
%%\end{lstlisting}
This is problematic, for all of the reasons described above: it requires explicit conversions between representations, interference issues with string syntax, correctness problems, performance overhead and security issues.

@@

So although languages that allow users to introduce new syntax from within libraries appear to hold promise for the reasons described above, providing this form of extensibility is non-trivial because there is no longer a central designer making decisions about such ambiguities. In most existing related work, the burden of resolving ambiguities falls to the clients using conflicting extensions. For example, SugarJ \cite{erdweg2011sugarj} and other extensible languages generated by Sugar* \cite{erdweg2013framework} allow providers to extend the base syntax of the host language with new forms, like set and map literals. These new forms are imported transitively throughout a program. To resolve syntactic ambiguities that arise, clients must manually augment the composed grammar with new rules that allow them to choose the correct interpretation explicitly. This is both difficult to do, requiring a reasonably thorough understanding of the underlying parser technology (in Sugar*, generalized LR parsing) and increases the cognitive load of using the conflicting notations (e.g. both sets and dictionaries) in the same file. These kinds of conflicts occur in a variety of circumstances: HTML and XML, different variants of SQL, JSON literals and dictionaries, or simply different implementations (``desugarings'') of the same specialized syntax (e.g. two regular expression engines) can all cause problems that may be difficult to anticipate.

In this work, we will describe an alternative parsing strategy that avoids these problems by shifting responsibility for parsing certain \emph{generic literal forms} into the typechecker. The typechecker, in turn, defers responsibility to user-defined types, by treating the body of the literal as a term of the   \emph{type-specific language (TSL)} associated with the type it is being checked against. The TSL is responsible for rewriting this term to ultimately use only general-purpose notation. This strategy avoids the problem of conflicting syntax, because neither the base language nor TSLs are ever extended directly. It also permits semantic flexibility -- the meaning of a form like \verb|{ }| can differ depending  on its type, so it is safe to use it for empty sets, dictionaries and other data structures, like JSON literals. This frees these common notations from being tied to the variant of a  data structure built into a language's standard library, which sometimes does not provide the exact semantics that a programmer needs (for example, Python dictionaries do not preserve key insertion order).
%\begin{lstlisting}
%let empty_set : Set = { }
%let empty_dict : Dict = { }
%let empty_json : JSON = { }
%\end{lstlisting}
\subsection{Wyvern}
We develop our work as a variant of a new programming language being developed by our group called Wyvern \cite{Nistor:2013:WST:2489828.2489830}. To allow us to focus on the essence of our proposal, the variant of Wyvern we will describe in this thesis is simpler than the variant previously described: it is purely functional (there are no effects other than non-termination) and it does not enforce a uniform access principle for objects (fields can be accessed directly). Objects are thus a simple variant of records equipped with simple methods (functions that are automatically given a self-reference) for convenience. We also add recursive labeled sum types, which we call \emph{case types}, that are quite similar to datatypes in ML. One can refer to the version of the language described in this thesis as \emph{TSL Wyvern}.

\subsection{Example: Web Search}
We begin in Fig. \ref{f-example} with an example showing several different TSLs being used to define a fragment of a web application showing search results from a database. Note that for clarity of presentation, we color each character  according to the TSL it is governed by.
\begin{figure}[t]
\begin{lstlisting}
let imageBase : URL = <SURLimages.example.comEURL>
let bgImage : URL = <SURL%EURLimageBaseSURL%/background.pngEURL>
new : SearchServer
  def resultsFor(searchQuery : String, page : Nat) : Unit =
    serve(~) (* serve : HTML -> Unit *)
SHTML      :html
        :head
          :title Search Results
          :style EHTML~
SCSS            body { background-image: url(%ECSSbgImageSCSS%) }
            #search { background-color: %ECSS`SCOLOR#aabbccECOLOR`.darken(SPCT20pctEPCT)SCSS% }
ECSSSHTML        :body
          :h1 Results for {EHTMLsearchQuerySHTML}
          :div[id="search"]
            Search again: {EHTMLSearchBox($\texttt{"}$SSTRGo!ESTR$\texttt{"}$)SHTML}
          {EHTML (* fmt_results : DB * SQLQuery * Nat * Nat -> HTML *)
            fmt_results(db, ~, SNAT10ENDNAT, page)
              SSQLSELECT * FROM products WHERE {ESQLsearchQuerySSQL} in titleESQL
          SHTML}
\end{lstlisting}
\vspace{-8px}
\caption{Wyvern Example with Multiple TSLs}
\label{f-example}
\vspace{-10px}
\end{figure}
\subsection{Inline Literals}
\begin{figure}[t]
\begin{lstlisting}
<SURLliteral body here, <inner angle brackets> must be balancedEURL>
{SURLliteral body here, {inner braces} must be balancedEURL}
[SURLliteral body here, [inner brackets] must be balancedEURL]
`SURLliteral body here, ``inner backticks`` must be doubledEURL`
$\texttt{'}$SURLliteral body here, ''inner single quotes'' must be doubledEURL$\texttt{'}$
$\texttt{"}$SURLliteral body here, ""inner double quotes"" must be doubledEURL$\texttt{"}$
SURL12xyzEURL (* no delimiters necessary for number literals; suffix optional *)
\end{lstlisting}
\vspace{-8px}
\caption{Inline Generic Literal Forms}
\vspace{-10px}
\label{f-delims}
\end{figure}
Our first TSL appears on the right-hand side of the variable binding on line 1. The variable \li{imageBase} is annotated with its type, \li{URL} (not shown). This is an {object type} declaring several fields representing the components of a URL: its protocol, domain name, port, path and so on. We could have created a value of type \li{URL} using general-purpose notation:
\begin{lstlisting}
let imageBase : URL = new
  val protocol : URLFrag = $\text{"}$SUShttpEUS$\texttt{"}$
  val subdomain : URLFrag = $\texttt{"}$SUSimagesEUS$\texttt{"}$
  (* ... *)
\end{lstlisting}
%  val domain : URLString = $\texttt{"}$SSTRexampleESTR$\texttt{"}$
This is tedious. Instead, because the \li{URL} type has a TSL associated with it, we can instead introduce precisely this value using conventional notation for URLs by placing it in the \emph{body} of a \emph{generic literal}, \li{<SURLimages.example.comEURL>}. Any other delimited form in Fig. \ref{f-delims} could equivalently be used if the constraints shown are obeyed. The type annotation on \li{imageBase} implies that this literal's \emph{expected type} is \li{URL}, so the {body} of the literal (the characters between the angle brackets, in blue) will be governed by the \li{URL} TSL during the typechecking phase. This TSL will parse the body ({at compile-time}) to produce a Wyvern abstract syntax tree (AST) that explicitly instantiates a new object of type \li{URL} using general-purpose notation.

In addition to supporting conventional notation for URLs, this TSL supports \emph{typed interpolation}
of another Wyvern expression of type \li{URL} to form a larger URL. The interpolated term is delimited by percent signs,
as seen on line 2 of Fig. \ref{f-example}. The TSL parses code between percent signs  as a Wyvern expression with expected type \li{URL}, using its AST to construct an AST for the expression as a whole. String interpolation 
never occurs. Note that the delimiters that are used to go from Wyvern to a TSL are controlled by Wyvern (those in Fig. \ref{f-delims}) while the TSL controls how to return to Wyvern.

%\subsection{General-Purpose Notation for Object and Case Types}
%The general-purpose introductory form for object types is \li{new}. This form is a syntactic \emph{forward reference} to the layout-delimited block of {definitions} beginning on the line immediately after the line \li{new} appears in (line 4 in this case), and ending when the indentation level has returned to the baseline, or when the file ends (after line 19 in this case). An object in TSL Wyvern can contain methods, introduced using \li{def}, and fields, introduced using \li{val}. Here we have just a single method, \li{serve_results} taking two arguments. Object types in TSL Wyvern are simple structural interfaces that constrain the signatures of fields and methods. The \emph{type ascription} around \li{new} checks that the object being introduced satisfies the signature of \li{SearchServer} (not shown).\todo{discuss case types}
\subsection{Layout-Delimited Literals}

\begin{figure}
\begin{lstlisting}[escapechar=$]
casetype HTML 
    Empty of Unit
  | Text of String
  | Seq of HTML * HTML 
  | BodyElement of Attributes * HTML
  | StyleElement  of Attributes * CSS
  | ...
  metadata = new : HasTSL
    val parser : Parser = ~
SGRM      start -> ':body'= start>
        EGRMfn child:Exp => `SQTHTML.BodyElement(([], %EQTchildSQT%))EQT`
SGRM      start -> ':style'= EXP>
        EGRMfn e:Exp => `SQTHTML.StyleElement(([], %EQTeSQT%))EQT`
SGRM      start -> '{'= EXP '}'=EGRM
        fn e:Exp => `SQT%EQTeSQT% : HTMLEQT`
\end{lstlisting}
\vspace{-8px}
\caption{A Wyvern case type with an associated TSL.}
\label{f-htmltype}
\end{figure}
On line 5 of Fig. \ref{f-example}, we see a call to a function \li{serve}, not shown, which has type \li{HTML -> Unit}. Here, \li{HTML} is a user-defined \emph{case type}, having cases for each HTML tag as well as some other structures, like text nodes and sequencing. Declarations  of some of these cases can be seen on lines 2-6 of Fig. \ref{f-htmltype} (note that TSL Wyvern also includes simple product types for convenience, written \li{T1 * T2}). We could again use Wyvern's general-purpose introductory form for case types, e.g. \li{HTML.BodyElement((attrs, child))} ({unlike in ML, in Wyvern we must explicitly qualify constructors with the case type they are part of when they are used}. This is largely to make our formal semantics simpler and for clarity of presentation.) But, as discussed above, using this syntax can be inconvenient and  cognitively demanding. Thus, we associate a TSL with \li{HTML} that provides a simplified notation for writing HTML, shown being used on lines 6-20 of Fig. \ref{f-example}. This literal body is layout-delimited, rather than delimited by explicit tokens as in Fig. \ref{f-delims}, and introduced by a form of \emph{forward reference}, written \li{~} (``tilde''), on the previous line. Because the forward reference occurs in a position where the expected type is \li{HTML}, the literal body is governed by that type's TSL. The forward reference will be replaced by the general-purpose term, of type \li{HTML}, generated by the TSL during typechecking.
\subsection{Implementing a TSL}

>>>>

% !TEX root = omar-thesis-proposal.tex
\section{Type-Specific Languages}\label{aparsing}
General-purpose constructs like those built into modern languages are, semantically, quite powerful. By using a general-purpose construct to define a data structure, one immediately benefits from a body of useful operators, established reasoning principles, well-optimized implementations and tool support. For example, lists can be encoded using a more general mechanism for defining polymorphic recursive sum types. Intuitively, a list can either be empty, or be broken down into a \emph{head} element and a \emph{tail}, another list. In an ML-like language, the polymorphic list type would be defined:
\begin{lstlisting}[numbers=none]
datatype 'a list = Nil | Cons of 'a * 'a list
\end{lstlisting}
By defining lists in this way, we can immediately reason about list values by structural induction and examine them by pattern matching. %The programmer only needs to provide an encoding of the structure of lists; the semantics and implementation are filled in by the general-purpose abstraction mechanism.
While this is quite useful, the associated general-purpose notation is not. For example, few would claim that writing a list of numbers as a sequence of \li{Cons} cells is convenient:
\begin{lstlisting}[numbers=none]
Cons(1, Cons(2, Cons(3, Cons(4, Nil))))
\end{lstlisting}

Because lists are a common data structure, many languages include \emph{literal notations} for introducing them, e.g. \li{[1, 2, 3, 4]}. This notation is semantically equivalent to the general-purpose notation shown above, but brings cognitive benefits by drawing attention to the content of the list, rather than the nature of the encoding. Using terminology from Green's cognitive dimensions of notations \cite{green1996usability}, it is more \emph{terse}, \emph{visible} and \emph{maps more closely} to the intuitive notion of a list. Stoy, in writing about the value of good notation, says \cite{stoybook}:
\begin{quote}A good notation thus conceals much of the inner workings behind suitable abbreviations, while allowing us to consider it in more detail if we require: matrix and tensor notations provide further good examples of this. It may be summed up in the saying: 'A notation is important for what it leaves out.'\end{quote}

Although list, number and string literals are nearly ubiquitous features of modern languages, some languages also  provide specialized notation for other common data structures (like maps, sets, vectors and matrices), data formats (like XML and JSON), query languages (like regular expressions and SQL), markup languages (like HTML and Markdown) and many other types of data. For example, a language with built-in notation for HTML and SQL, supporting type-safe \emph{interpolation}\todo{splicing?} via curly braces, might define:
\begin{lstlisting}
let webpage : HTML = SHTML<html><body><h1>Results for {EHTMLkeywordSHTML}</h1>
  <ul id="results">{EHTMLto_list_items(query(db, 
    SSQLSELECT title, snippet FROM products WHERE {ESQLkeywordSSQL} in titleESQL)SHTML}
  </ul></body></html>EHTML
\end{lstlisting}
as shorthand for:
\begin{lstlisting}
let webpage : HTML = HTMLElement(Dict.empty(), [BodyElement(Dict.empty(),
  [H1Element(Dict.empty(), [TextNode(concat($\texttt{"}$SSTRResults for $\texttt{"}$ESTR, keyword))]), 
  ULElement((Dict.add Dict.empty() ($\texttt{"}$SSTRid$\texttt{"}$ESTR, $\texttt{"}$SSTRresults$\texttt{"}$ESTR)), to_list_items(query(db, 
    SelectStmt([$\texttt{"}$SSTRtitle$\texttt{"}$ESTR, $\texttt{"}$SSTRsnippet$\texttt{"}$ESTR], $\texttt{"}$SSTRproducts$\texttt{"}$ESTR, 
      [WhereClause(InPredicate(StringLit(keyword), $\texttt{"}$SSTRtitle$\texttt{"}$ESTR))]))))]]]
\end{lstlisting}

When a specialized notation like this is not available, but the equivalent general-purpose notation is too cognitively demanding for comfort, developers typically turn to run-time mechanisms to make constructing data structures more convenient. Among the most common strategies across language paradigms in these situations is to simply use a string representation that is parsed at run-time.% Developers across language paradigms frequently write examples like the above as:
\begin{lstlisting}
let webpage : HTML = parse_html($\texttt{"}$SSTR<html><body><h1>Results for ESTR$\texttt{"}$ + keyword + $\texttt{"}$SSTR</h1>
  <ul id=\$\texttt{\color{cyan}"}$results\$\texttt{\color{cyan}"}$>$\texttt{"}$ESTR + to_string(to_list_items(query(db, parse_sql(
  	$\texttt{"}$SSTRSELECT title, snippet FROM products WHERE '$\texttt{" + keyword + "}$' in title$\texttt{"}$ESTR)))) + 
  $\texttt{"}$SSTR</ul></body></html>$\texttt{")}$
\end{lstlisting}

Though recovering much of the notational convenience of the literal version, it is still more awkward to write, requiring explicit conversions to and from structured representations (\li{parse_html} and \li{to_string}, respectively) and escaping when the syntax of the language interferes with the syntax of string literals (line 2). Code like this also causes a number of problems beyond cognitive load. Because parsing occurs at run-time, syntax errors will not be discovered statically, causing potential problems in production scenarios. Run-time parsing also incurs performance overhead, particularly relevant when code like this is executed often (as on a heavily-trafficked website). But the most serious issue with this code is that it is fundamentally insecure: it is vulnerable to cross-site scripting attacks (line 1) and SQL injection attacks (line 3). For example, if a user entered the keyword \li{'; DROP TABLE products --}, the entire product database could be erased. These attack vectors are considered to be two of the most serious security threats on the web today \cite{owasp2013}. Although developers are cautioned to sanitize their input, it can be difficult to verify that this was done correctly throughout a codebase. The best way to avoid these problems today is to avoid strings and insist on structured representations, despite the inconvenient notation.

Unfortunately, situations like this, where maintaining strong correctness, performance and security guarantees entails significant syntactic overhead, causing developers to turn to worse solutions that are more convenient, are quite common. To emphasize this, let us return to our running example of pattern literals. A small regular expression like \verb!(\d\d):(\d\d)\w*((am)|(pm))! might be written using general-purpose notation as:
\begin{lstlisting}
Seq(Group(Seq(Digit, Digit), Seq(Char($\texttt{"}$SSTR:ESTR$\texttt{"}$), Seq(Group(Seq(Digit, Digit)), 
  Seq(ZeroOrMore(Whitespace), Group(Or(Group(Seq(Char($\texttt{"}$SSTRaESTR$\texttt{"}$), Char($\texttt{"}$SSTRmESTR$\texttt{"}$))), 
  Group(Seq(Char($\texttt{"}$SSTRpESTR$\texttt{"}$), Char($\texttt{"}$SSTRmESTR$\texttt{"}$))))))))))
\end{lstlisting}
This is clearly more cognitively demanding, both when writing the regular expression and when reading it. Among the most common strategies in these situations, for users of any kind of language, is again to simply use a string representation that is parsed at run-time.
\begin{lstlisting}
rx_from_str($\texttt{"}$SSTR(\\d\\d):(\\d\\d)\\w*((am)|(pm))ESTR$\texttt{"}$)
\end{lstlisting}
%%
%%For example, in languages without SQL literals, developers can implement a builder pattern:
%%\begin{lstlisting}
%%new SQLQuery().SELECT("*").FROM("table").WHERE("username").Eq(username)
%%\end{lstlisting}
This is problematic, for all of the reasons described above: it requires explicit conversions between representations, interference issues with string syntax, correctness problems, performance overhead and security issues.

@@

So although languages that allow users to introduce new syntax from within libraries appear to hold promise for the reasons described above, providing this form of extensibility is non-trivial because there is no longer a central designer making decisions about such ambiguities. In most existing related work, the burden of resolving ambiguities falls to clients who have the misfortune of importing conflicting extensions. For example, SugarJ \cite{erdweg2011sugarj} and other extensible languages generated by Sugar* \cite{erdweg2013framework} allow providers to extend the base syntax of the host language with new forms, like set and map literals. These new forms are imported transitively throughout a program. To resolve syntactic ambiguities that arise, clients must manually augment the composed grammar with new rules that allow them to choose the correct interpretation explicitly. This is both difficult to do, requiring a reasonably thorough understanding of the underlying parser technology (in Sugar*, generalized LR parsing) and increases the cognitive load of using the conflicting notations (e.g. both sets and dictionaries) in the same file because disambiguation tokens must be used. These kinds of conflicts occur in a variety of circumstances: HTML and XML, different variants of SQL, JSON literals and dictionaries, or simply different implementations (``desugarings'') of the same specialized syntax (e.g. two regular expression engines) can all cause problems that may be difficult to anticipate.

In this work, we will describe an alternative parsing strategy that avoids these problems by shifting responsibility for parsing certain \emph{generic literal forms} into the typechecker. The typechecker, in turn, defers responsibility to user-defined types, by treating the body of the literal as a term of the   \emph{type-specific language (TSL)} associated with the type it is being checked against. The TSL is responsible for rewriting this term to ultimately use only general-purpose notation. This strategy avoids the problem of conflicting syntax, because neither the base language nor TSLs are ever extended directly. It also permits semantic flexibility -- the meaning of a form like \verb|{ }| can differ depending  on its type, so it is safe to use it for empty sets, maps and other data structures, like JSON literals. This frees these common notations from being tied to the variant of a  data structure built into a language's standard library, which sometimes does not provide the exact semantics that a programmer needs (for example, Python dictionaries do not preserve key insertion order).
\begin{figure}[t]
\begin{lstlisting}
let imageBase : URL = <SURLimages.example.comEURL>
let bgImage : URL = <SURL%EURLimageBaseSURL%/background.pngEURL>
new : SearchServer
  def resultsFor(searchQuery : String, page : Nat) : Unit =
    serve(~) (* serve : HTML -> Unit *)
SHTML      :html
        :head
          :title Search Results
          :style EHTML~
SCSS            body { background-image: url(%ECSSbgImageSCSS%) }
            #search { background-color: %ECSS`SCOLOR#aabbccECOLOR`SCSS% }
ECSSSHTML        :body
          :h1 Results for :{EHTMLsearchQuerySHTML}
          :div[id="search"]
            Search again: :{EHTMLSearchBox($\texttt{"}$SSTRGo!ESTR$\texttt{"}$)SHTML}
          :wyvEHTML (* fmt_results : DB * SQLQuery * Nat * Nat -> HTML *)
            fmt_results(db, ~, SNAT10ENDNAT, page)
              SSQLSELECT * FROM products WHERE {ESQLsearchQuerySSQL} in titleESQL
\end{lstlisting}
\vspace{-8px}
\caption{Wyvern Example with Multiple TSLs}
\label{f-example}
\vspace{-10px}
\end{figure}
%\begin{lstlisting}
%let empty_set : Set = { }
%let empty_dict : Dict = { }
%let empty_json : JSON = { }
%\end{lstlisting}
\subsection{Wyvern}
We develop our work as a variant of a new programming language being developed by our group called Wyvern \cite{Nistor:2013:WST:2489828.2489830}. To allow us to focus on the essence of our proposal, the variant of Wyvern we will describe in this thesis is simpler than the variant previously described: it is purely functional (there are no effects other than non-termination) and it does not enforce a uniform access principle for objects (fields can be accessed directly). Objects can thus be thought of as functional records equipped with simple methods (functions that are automatically given a self-reference) for convenience. We also add recursive labeled sum types, which we call \emph{case types}, that are quite similar to datatypes in ML. One can refer to the version of the language described in this thesis as \emph{TSL Wyvern}.

\subsection{Example: Web Search}
We begin in Fig. \ref{f-example} with an example showing several different TSLs being used to define a fragment of a web application showing search results from a database. Note that for clarity of presentation, we color each character  according to the TSL it is governed by.% Black represents the base language.

\subsection{Inline Literals}
\begin{figure}[t]
\begin{lstlisting}[numbers=none]
<SURLliteral body here, <inner angle brackets> must be balancedEURL>
{SURLliteral body here, {inner braces} must be balancedEURL}
[SURLliteral body here, [inner brackets] must be balancedEURL]
`SURLliteral body here, ``inner backticks`` must be doubledEURL`
$\texttt{'}$SURLliteral body here, ''inner single quotes'' must be doubledEURL$\texttt{'}$
$\texttt{"}$SURLliteral body here, ""inner double quotes"" must be doubledEURL$\texttt{"}$
SURL12xyzEURL (* no delimiters necessary for number literals; suffix optional *)
\end{lstlisting}
\vspace{-8px}
\caption{Inline Generic Literal Forms}
\vspace{-10px}
\label{f-delims}
\end{figure}
Our first TSL appears on the right-hand side of the variable binding on line 1. The variable \li{imageBase} is annotated with its type, \li{URL}. This is an {object type} declaring several fields representing the components of a URL: its protocol, domain name, port, path and so on (not shown). We could have created a value of type \li{URL} using general-purpose notation:
\begin{lstlisting}
let imageBase : URL = new
  val protocol = $\text{"}$SUShttpEUS$\texttt{"}$
  val subdomain = $\texttt{"}$SUSimagesEUS$\texttt{"}$
  (* ... *)
\end{lstlisting}
%  val domain : URLString = $\texttt{"}$SSTRexampleESTR$\texttt{"}$
This is tedious. Because the \li{URL} type has a TSL associated with it, we can instead introduce precisely this value using conventional notation for URLs by placing it in the \emph{body} of a \emph{generic literal}, \li{<SURLimages.example.comEURL>}. Any other delimited form in Fig. \ref{f-delims} could equivalently be used if the constraints shown are obeyed. The type annotation on \li{imageBase} implies that this literal's \emph{expected type} is \li{URL}, so the {body} of the literal (the characters between the angle brackets, in blue) will be governed by the \li{URL} TSL during the typechecking phase. This TSL will parse the body ({at compile-time}) to produce a Wyvern abstract syntax tree (AST) that explicitly instantiates a new object of type \li{URL} using general-purpose notation as if the above had been written directly. We will return to how this works shortly.

In addition to supporting conventional notation for URLs, this TSL supports \emph{interpolating}
another Wyvern expression of type \li{URL} to form a larger URL. The interpolated term is delimited by percent signs,
as seen on line 2 of Fig. \ref{f-example}. The TSL parses code between percent signs  as a Wyvern expression, using its AST to construct an AST for the expression as a whole. String concatenation  
never occurs. Note that the delimiters used to go from Wyvern to a TSL are controlled by Wyvern  while the TSL controls how to return to Wyvern. 
%\subsection{General-Purpose Notation for Object and Case Types}
%The general-purpose introductory form for object types is \li{new}. This form is a syntactic \emph{forward reference} to the layout-delimited block of {definitions} beginning on the line immediately after the line \li{new} appears in (line 4 in this case), and ending when the indentation level has returned to the baseline, or when the file ends (after line 19 in this case). An object in TSL Wyvern can contain methods, introduced using \li{def}, and fields, introduced using \li{val}. Here we have just a single method, \li{serve_results} taking two arguments. Object types in TSL Wyvern are simple structural interfaces that constrain the signatures of fields and methods. The \emph{type ascription} around \li{new} checks that the object being introduced satisfies the signature of \li{SearchServer} (not shown).\todo{discuss case types}
\subsection{Layout-Delimited Literals}

On line 5 of Fig. \ref{f-example}, we see a call to a function \li{serve} (not shown) which has type \li{HTML -> Unit}. Here, \li{HTML} is a user-defined \emph{case type}, having cases for each HTML tag as well as some other structures, like text nodes and sequencing. Declarations  of some of these cases can be seen on lines 2-6 of Fig. \ref{f-htmltype} (note that TSL Wyvern also includes simple product types for convenience, written \li{T1 * T2}). We could again use Wyvern's general-purpose introductory form for case types, e.g. \li{HTML.BodyElement((attrs, child))} ({unlike in ML, in Wyvern we must explicitly qualify constructors with the case type they are part of when they are used}. This is largely to make our formal semantics simpler and for clarity of presentation.) But, as discussed above, using this syntax can be inconvenient and  cognitively demanding. Thus, we associate a TSL with \li{HTML} that provides a simplified notation for writing HTML, shown being used on lines 6-18 of Fig. \ref{f-example}. This literal body is layout-delimited, rather than delimited by explicit tokens as in Fig. \ref{f-delims}, and introduced by a form of \emph{forward reference}, written \li{~} (``tilde''), on the previous line. Because the forward reference occurs in a position where the expected type is \li{HTML}, the literal body is governed by that type's TSL. The forward reference will be replaced by the general-purpose term, of type \li{HTML}, generated by the TSL during typechecking. The use of layout imposes no syntactic constraints on the body, while using an inline form does (Fig. \ref{f-delims}). For HTML, this is quite useful, as all of the inline forms impose constrains that would cause conflict with some valid piece of HTML.
\subsection{Implementing a TSL}
\begin{figure}
\begin{lstlisting}[escapechar=$]
casetype HTML 
    Empty of Unit
  | Text of String
  | Seq of HTML * HTML 
  | BodyElement of Attributes * HTML
  | StyleElement  of Attributes * CSS
  | ...
  metadata = new : HasTSL
    val parser : Parser = ~
SGRM      start -> ':body'= start>
        EGRMfn child:Exp => `SQTHTML.BodyElement(([], %EQTchildSQT%))EQT`
SGRM      start -> ':style'= EXP>
        EGRMfn e:Exp => `SQTHTML.StyleElement(([], %EQTeSQT%))EQT`
SGRM      start -> '{'= EXP '}'=EGRM
        fn e:Exp => `SQT%EQTeSQT% : HTMLEQT`
\end{lstlisting}
\vspace{-8px}
\caption{A Wyvern case type with an associated TSL.}
\label{f-htmltype}
\end{figure}
\begin{figure}[t]
\begin{subfigure}[t]{.53\textwidth}
\begin{lstlisting}
objtype HasTSL
  val parser : Parser
objtype Parser                          
  def parse(ts : TokenStream) : (Exp * 
    TokenStream)
  metadata = new : HasTSL
    val parser : Parser = (* parser generator *)      
casetype Type
    Var of ID
  | Arrow of Type * Type
